---
output: 
  pdf_document:
    highlight: tango
    keep_tex: no
sansfont: Times New Roman
fontsize: 10.9pt
geometry: margin = 1in
header-includes:
  - \usepackage{titling}
  - \usepackage{graphicx}
  - \usepackage{fancyhdr}
  - \usepackage{background}
  - \usepackage{hyperref}
  - \usepackage{caption}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{wrapfig}
  - \usepackage{graphicx}
  - \usepackage{ragged2e}
  - \captionsetup[figure]{labelformat=empty}
  - \captionsetup[table]{labelformat=empty}
  - \pagestyle{fancy}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \fancyhf{}
  - \fancyhead[L]{\ifnum\value{page} < 2 {\colorbox{black}{\parbox[b][2.5cm][c]{16.3cm}{\textcolor{white}{Eastern Forest Working Group\\ Inventory and Monitoring Program\\ \hspace{1cm}\\Forest Health Monitoring\\ \hspace{1cm}}}}}\else {Eastern Forest Working Group}\small \fi}
  - \chead[C]{\ifnum\value{page} < 2 {\textcolor{white}{\hspace{5.7cm}National Park Service\\ \hspace{7.2cm}U.S. Department of the Interior\\ \vspace{1.1cm}}} \else \fi}
  - \fancyhead[R]{\ifnum\value{page} < 2 {\includegraphics[height=.95in,width=.95in]{inputs/NPS_logo.png}} \else {\today}\small \fi}
  - \setlength{\footskip}{34pt}
  - \fancyfoot[C]{\thepage\footnotesize}
  - \rfoot{\ifnum\value{page} > 1 \href{https://schoodicinstitute.org}{\includegraphics[height=\footskip]{inputs/Schoodic_Stacked.jpeg}} \href{https://www.nps.gov/index.htm}{\includegraphics[height=\footskip]{inputs/NPS_logo.png}} \else \fi}
  - \backgroundsetup{scale=1,opacity=0.5,angle=0,placement=bottom,contents={\ifnum\value{page} > 1 {\includegraphics[width=8.5in]{inputs/background2.jpg}} \else \fi}}
  - \pretitle{\begin{center}\LARGE\bfseries\vspace{.5cm}}
  - \posttitle{\end{center}}
  - \setlength{\abovecaptionskip}{3pt}
  - \setlength{\belowcaptionskip}{0pt}
  - \setlength{\intextsep}{0pt}
params:
  park: park
---

\newenvironment{specwrapfigure}
 {%
  \setlength{\columnsep}{15pt}%
  \wrapfloat{figure}%
 }
 {\endwrapfloat}
 
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.pos = "!h")
```

```{r imports, include = FALSE, results = 'hide'}
library(tidyverse)
library(forestTrends)
library(png)
library(knitr)
library(cowplot)
library(magick)
library(kableExtra)

dens_df <- read.csv("../data/EFWG_full_dataset_20211101.csv")

park_info <- read.csv("../data/park_names.csv") %>% 
  filter(park_full == park_name)

park_data <- read.csv("../results/20220325/final_park_results.csv") %>% 
  filter(park == paste(park_info$park)) %>% 
  filter(labels == "Seedling Density" & metgrp == "Regen. Status") %>% 
  mutate(park_reggrp = ifelse(park_reggrp == "Sec.", "Secure", park_reggrp),
         metgrp = str_replace(metgrp, "Regen.", "Regeneration"))

park_table <- read.csv("../results/20220325/final_park_results.csv") %>% 
  filter(park == paste(park_info$park)) %>% 
  mutate(park_reggrp = ifelse(park_reggrp == "Sec.", "Secure", park_reggrp),
         metgrp = str_replace(metgrp, "Regen.", "Regeneration"),
         sign = str_replace(sign, "nonsign", "no supported trend"),
         sign = str_replace(sign, "notmod", "not modeled"),
         sign = str_replace(sign, "signdec_bad", "decrease"),
         sign = str_replace(sign, "signdec_good", "decrease"),
         sign = str_replace(sign, "signinc_bad", "increase"),
         sign = str_replace(sign, "signinc_good", "increase"))

park_text <- read.csv("../data/parks_mgt_20220831.csv") %>% 
  rename_with(tolower) %>% 
  filter(park == paste(park_info$park))
```

``` {r text, include = FALSE, results = 'hide'}

# If statements for text
mgtst <- if (park_text$current.active.deer.management == "yes" & 
    park_text$reccomend.deer.management == "yes" &
    park_text$thinning...canopy.gaps == "no" &
    park_text$prescribed.burning == "no") {
  
  paste("We also recommend that there is continued effort to manage deer populations to help avoid forest loss in the park.")
    
  } else if (park_text$current.active.deer.management == "no" & 
             park_text$reccomend.deer.management == "yes" &
             park_text$thinning...canopy.gaps == "no" &
             park_text$prescribed.burning == "no") {
    
    paste("We also recommend that long-term deer management strategies be implemented to help avoid forest loss in the park.")
    
  } else if (park_text$current.active.deer.management == "no" & 
             park_text$reccomend.deer.management == "yes" &
             park_text$thinning...canopy.gaps == "no" &
             park_text$prescribed.burning == "yes") {
    
    paste("We also recommend that long-term deer management strategies be implemented, as well as prescribed burning, to help avoid forest loss in the park.")
    
  } else if (park_text$current.active.deer.management == "no" & 
             park_text$reccomend.deer.management == "no" &
             park_text$thinning...canopy.gaps == "no" &
             park_text$prescribed.burning == "no") {
    
    paste("Although we do not recommend any other management stategies at this time, conditions are changing rapidly across parks and continued monitoring and reassessment of management actions is necessary to avoid future forest loss in the park.")
    
  } else if (park_text$current.active.deer.management == "no" & 
             park_text$reccomend.deer.management == "no" &
             park_text$thinning...canopy.gaps == "yes" &
             park_text$prescribed.burning == "yes") {
    
    paste("We also recommend the assessment and application of other management actions including, thinning, canopy gap management, and prescribed burning, to help avoid forest loss in the park.")
    
  }


notable <- if (park_text$notable.findings == "none") {
  
  paste("")
  
} else if (park_text$notable.findings == "forest loss projected from lack of young trees") {
  
  paste(" ", park_info$park, " has projected forest loss due to lack of young trees.")
  
} else if (park_text$notable.findings == "increase exotic trees/sap/seed") {
  
  paste(" ", park_info$park, " shows an increase in exotic trees, saplings, and/or seedlings.")
  
} else if (park_text$notable.findings == "increase exotic seedlings") {
  
  paste(" ", park_info$park, " shows an increase in exotic seedlings.")
  
} else if (park_text$notable.findings == "decline in native seedlings") {
  
  paste(" ", park_info$park, " shows declines in native seedlings.")
  
} else if (park_text$notable.findings == "increase native canopy saplings") {
  
  paste(" ", park_info$park, " shows an increase in native canopy saplings.")
  
} else if (park_text$notable.findings == "regeneration mismatch ") {
  
  paste0(" ", park_info$park, " is experiencing regeneration mismatch between the forest canopy layer and seedling layer due to an abundance of regeneration from species which do not form an overstory canopy. In this park, the mismatch is occuring due to ", park_text$x, ".")
  
}
```

\hfill\break
\hfill\break
\begin{center}\huge\bf {The Current Forest Regeneration Status for `r paste(park_info$park_full)` is: `r paste(park_data$park_reggrp)`}\end{center}

\begin{specwrapfigure}{r}{0.7\textwidth}
  \centering
    \includegraphics[width=\linewidth]{inputs/Region1_manuscript.jpg}
  \caption{{\bf Figure 1}. Map of parks included in regional regeneration project and forest regeneration status.}
\end{specwrapfigure}

\begin{flushleft}\large {}\end{flushleft}

\begin{flushleft}\large\bf {Introduction}\end{flushleft}
`r paste(park_info$park_full)` (`r paste(park_info$park)`) is located in the `r ifelse(park_info$Network == "NETN", paste("Northeast Temperate"), ifelse(park_info$Network == "NCRN", paste("National Capital Region"), ifelse(park_info$Network == "NCBN", paste("Northeast Coastal and Barrier"), ifelse(park_info$Network == "MIDN", paste("Mid-Atlantic"), paste("Eastern Rivers and Mountains")))))` `r paste0( '(', park_info$Network, ')')` Inventory and Monitoring Network (Figure 1). This brief summarizes the notable findings from the Eastern Forest Working Group Forest Health Monitoring project, and recommends management actions to prevent the loss of forests in `r paste(park_info$park)`. For more detailed information see the source publication in the 'Resources' section below.

\begin{flushleft}\small {}\end{flushleft}

\begin{flushleft}\large\bf {Main Findings}\end{flushleft}
On a four category scale (Secure, Insecure, Probable Failure, and Imminent Failure), `r paste(park_info$park)` has a forest regeneration status of \textbf{`r paste(park_data$park_reggrp)`}. This means that the park `r ifelse(park_data$park_reggrp == "Imminent Failure", paste("is experiencing severe regeneration failure, and is at risk of forest loss. In some cases, this forest loss may already be underway"), ifelse(park_data$park_reggrp == "Probable Failure", paste("is experiencing regeneration failure as indicated by lacking and/or declining sapling numbers, and low numbers of seedlings"), ifelse(park_data$park_reggrp == "Insecure", paste("currently lacks saplings, but has sufficient and often increasing seedling numbers, and overall lower deer browse impacts. This tends to be due to forest dynamics, such as succession"), paste("currently has abundant seedling and sapling numbers, and shows few concerning trends in regeneration metrics"))))`.`r paste(notable)``r ifelse(park_text$findings.2 != "none", paste0(" Additionally, we found that there is ", park_text$findings.2, " in the park."), paste(""))` The regeneration status for each of the 10 metrics examined in this study are described in Figure 2. Depicted in Figure 3 are the trends in tree, sapling, and seedling abundance by native canopy, other native species, and exotics. 

\centering\includegraphics[width=.7\linewidth]{`r paste0("../results/parkregen_plots/", park_info$park, ".png")`}
\captionof{figure}{{\bf Figure 2}. Regeneration status for each of the 10 metrics in the study.}



\centering\includegraphics[width=1\linewidth]{`r paste0("../results/20220325/park_plots/spp_grp/", park_info$park, "_spp_grp.png")`}
\captionof{figure}{{\bf Figure 3}. Trends in tree, sapling and seedling abundance by species group. Trends are based on change over time across three complete survey cycles: Cycle 1 (2008 – 2011), Cycle 2 (2012 – 2015), and Cycle 3 (2016 – 2019).}

\hfill\break

\begin{flushleft}\large\bf {Management Recommendations}\end{flushleft}
\begin{flushleft}\tiny {}\end{flushleft}
\begin{flushleft}\normalsize {We suggest that managers of `r paste(park_info$park)` continue invasive plant management actions, as is true for all of the eastern national parks in this study. `r paste(mgtst)` Protected forests such as those in eastern national parks need forest stewardship plans, along with the resources to implement them, that address sustained commitment to reducing deer browse impacts, managing invasive plants, and enhancing forest structural complexity and diversity to secure sufficient regeneration. `r ifelse(park_data$park_reggrp == "Imminent Failure" | park_data$park_reggrp == "Probable Failure", paste0("These efforts are especially important for ", park_info$park, " as the park is very likely to lose forest cover without intervention."), paste(""))` Please refer to the source publication for more information on management strategies, and reach out to the contact below for further assistance.

}\end{flushleft}

\hfill\break
\hfill\break

\begin{center}\large\bf {Resources}\end{center}
\begin{flushleft}\normalsize {{\bf Source Publication:}\hfill\break Miller K., Perles S., Schmit J.P., Matthews E., Weed A., Comiskey J., Marshall M., Nelson P., Fisichelli N. 202X. Forests in eastern national parks face widespread regeneration debt. Ecological Applications.}\end{flushleft}

\begin{flushleft}\normalsize {}\end{flushleft}

\begin{flushleft}\normalsize {\bf Contact:}\end{flushleft}
`r ifelse(park_info$Network == "NETN", paste("Kate Miller, NETN Quantitative Ecologist (kathryn_miller@nps.gov)"), ifelse(park_info$Network == "MIDN", paste("Kate Miller, MIDN Quantitative Ecologist (kathryn_miller@nps.gov)"), ifelse(park_info$Network == "NCBN", paste("Kate Miller, NCBN Quantitative Ecologist (kathryn_miller@nps.gov)"), ifelse(park_info$Network == "NCRN", paste("Elizabeth Matthews, NCRN Regional Program Manager (elizabeth_matthews@nps.gov)"), ifelse(park_info$Network == "ERMN", paste("Stephanie Perles, ERMN Plant Ecologist (stephanie_perles@nps.gov)"), paste("ERROR"))))))` \hfill\break

\begin{flushleft}\normalsize {{\bf Links:}\hfill\break
1. \href{https://www.nps.gov/articles/series.htm?id=B9C6B751-A7E3-F246-5628389A0B80A4F7}{Managing Resilient Forests Initiative for Eastern National Parks}}\end{flushleft}

